---
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
```
<p style="text-align: center"> 
# **NYPD Shooting Incident - Historic Data Analysis**
</p>

> This report analyzes the data relating to shooting incidents which have occurred in NYC boroughs between 2006 to 2020. This data is extracted and reviewed every quarter by the Office of Management Analysis and Planning, and will be used in an overall analysis of shootings in each borough.

## Importing libraries

> First, the libraries that are needed for this analysis are loaded. 

```{r libraries, message=FALSE, warning=FALSE}
library(lubridate)
library(tidyverse)
library(knitr)
library(incidence)
```

# Importing Data

> Data is read in from the .csv file from the DATA.GOV website

```{r}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv"
nyshootings <- read.csv(url)
colnames(nyshootings)
```

# Data Wrangling

> Now that our data has been read in, it needs to be cleaned up and any data that are not pertinent to this analysis can be removed. Because I'm only interested in shootings in the boroughs, I am removing data such as Longitude and Latitude. 

```{r wrangling}
nyshootings <- nyshootings %>% select(-c(Latitude, Longitude, Lon_Lat,
                                         INCIDENT_KEY, JURISDICTION_CODE))
nyshootings <- mutate(nyshootings, OCCUR_DATE = mdy(OCCUR_DATE))
summary(nyshootings)
```

# Data Analysis

> After cleaning up the data we can analyse the data. The first analysis will be the total number of shootings per borough.

```{r analysis}
borough_totals <- table(nyshootings$BORO)
borough_totals <- as.data.frame(borough_totals) %>%
rename(Borough = Var1, Frequency = Freq)
borough_totals$Percent <- round((borough_totals$Frequency / sum(borough_totals$Frequency)*100),2)
kable(borough_totals)

# This summary can be visualized in a histogram

ggplot(borough_totals, aes(x=Borough, y=Frequency, fill=Borough)) + geom_bar(stat="identity")
```

> The histogram shows that the frequency of shootings in Brooklyn is the highest, with Staten Island having the lowest incidence. But this is the total shootings from 2006 to 2020, what has this data looked over time? We can analyze the data further by taking shooting incidence from Brooklyn and Staten Island and plotting against the year.

# Brooklyn

```{r brooklyn}
BR <-subset(nyshootings, BORO=='BROOKLYN', select=c(BORO, OCCUR_DATE))
n <- 5

BR$YEAR <- substr(BR$OCCUR_DATE, nchar(BR$OCCUR_DATE) - n + 1, nchar(BR$OCCUR_DATE)) 

BR <- subset(BR, select = -c(OCCUR_DATE))

BROOKLYN <- table(BR$YEAR)
BROOKLYN <- as.data.frame(BROOKLYN)

names(BROOKLYN)[names(BROOKLYN) == "Var1"] <- "Year"
names(BROOKLYN)[names(BROOKLYN) == "Freq"] <- "Frequency"

ggplot(data=BROOKLYN, aes(x=Year, y=Frequency, group=1)) +
  geom_line(colour = "blue")+
  geom_point(colour = "blue")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Staten Island

```{r staten_island}
SI <-subset(nyshootings, BORO=='STATEN ISLAND', select=c(BORO, OCCUR_DATE))
n <- 5   

SI$YEAR <- substr(SI$OCCUR_DATE, nchar(SI$OCCUR_DATE) - n + 1, nchar(SI$OCCUR_DATE))

SI <- subset(SI, select = -c(OCCUR_DATE))

STATEN_ISLAND <- table(SI$YEAR)
STATEN_ISLAND <- as.data.frame(STATEN_ISLAND)

names(STATEN_ISLAND)[names(STATEN_ISLAND) == "Var1"] <- "Year"
names(STATEN_ISLAND)[names(STATEN_ISLAND) == "Freq"] <- "Frequency"

ggplot(data=STATEN_ISLAND, aes(x=Year, y=Frequency, group=1)) +
  geom_line(colour = "red")+
  geom_point(colour = "red")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Shootings Overall

> The plots from Brooklyn and Staten Island show an overall decline with a sharp increase in shootings in the year 2020. Since this is only in 2 of the boroughs, we can use a model to determine whether shootings in NYC as a whole have increased or decreased over time. 

```{r model, message=FALSE, warning=FALSE}
sh <- incidence (nyshootings$OCCUR_DATE)
sh_fit <- fit(sh)
plot(sh, fit = sh_fit, color = "blue")
```

# Conclusion

> This data set contained a lot of data which could be used for analysis, from victim and perpretrator demographics, daily/yearly shootings, and even the latitude and longitude of each shooting. This particular analysis focused on the overall totals for each of the boroughs listed as well as breaking down the totals for the boroughs with the most and least shootings: Brooklyn and Staten Island. 

> Overall, in 2020, there has been an increase in the frequency of shootings in both Brooklyn and Staten Island as seen in the spike of each graph. However using the model, we can see that the overall trend of shootings in NYC as a whole trended downward. This analysis doesn't take in to consideration any outside influences on these numbers, nor does it deal with the demographics of the victims and perpretrators, or whether the size or geography of each borough played a role. This could be potential sources of bias in the analysis. On a personal level, I have no connection to NYC and the boroughs listed, and chose my analysis based on the shooting data. Being an immigrant to America, my knowledge of these places is based on TV shows and news stories, which create a favorable vs. unfavourable image that could be a source of bias for future analysis. 